---
title: "US Senate Scraper"
author: "ODONNELL"
date: "April 27, 2019"
output: html_document
---

## The overall goal of this R file is to create a
## dataframe with a row for each year of every
## US senators 

## Then, we can easily aggregate senator information
## by year

## First, load the needed packages:

```{r eval=TRUE, warning=FALSE, message=FALSE}
# first, for scraping
library(rvest)
library(RCurl)

# second, for tidying
library(tidyr)
library(dplyr)
```


## Next, scrape the needed Wikipedia tables
## 25 tables, one for each letter (no 'X')

```{r warning=FALSE}
#wikipedia page with all senators
page <- "https://en.wikipedia.org/wiki/List_of_former_United_States_Senators"

#collect the senators
senators.a <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[3]') %>%
    html_table()

senators.b <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[4]') %>%
    html_table()

senators.c <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[5]') %>%
    html_table()

senators.d <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[6]') %>%
    html_table()

senators.e <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[7]') %>%
    html_table()

senators.f <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[8]') %>%
    html_table()

senators.g <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[9]') %>%
    html_table()

senators.h <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[10]') %>%
    html_table()

senators.i <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[11]') %>%
    html_table()

senators.j <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[12]') %>%
    html_table()

senators.k <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[13]') %>%
    html_table()

senators.l <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[14]') %>%
    html_table()

senators.m <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[15]') %>%
    html_table()

senators.n <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[16]') %>%
    html_table()

senators.o <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[17]') %>%
    html_table()

senators.p <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[18]') %>%
    html_table()

senators.q <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[19]') %>%
    html_table()

senators.r <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[20]') %>%
    html_table()

senators.s <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[21]') %>%
    html_table()

senators.t <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[22]') %>%
    html_table()

senators.u <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[23]') %>%
    html_table()

senators.v <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[24]') %>%
    html_table()

senators.w <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[25]') %>%
    html_table()

senators.y <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[26]') %>%
    html_table()

senators.z <- page %>% 
    html() %>%
    html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[27]') %>%
    html_table()
```

### Third, drop the unneeded 'Class' column
### ..this column name does not match among tables

```{r eval=TRUE}
senators.a[[1]] = senators.a[[1]][,-3]
senators.b[[1]] = senators.b[[1]][,-3]
senators.c[[1]] = senators.c[[1]][,-3]
senators.d[[1]] = senators.d[[1]][,-3]
senators.e[[1]] = senators.e[[1]][,-3]
senators.f[[1]] = senators.f[[1]][,-3]
senators.g[[1]] = senators.g[[1]][,-3]
senators.h[[1]] = senators.h[[1]][,-3]
senators.i[[1]] = senators.i[[1]][,-3]
senators.j[[1]] = senators.j[[1]][,-3]
senators.k[[1]] = senators.k[[1]][,-3]
senators.l[[1]] = senators.l[[1]][,-3]
senators.m[[1]] = senators.m[[1]][,-3]
senators.n[[1]] = senators.n[[1]][,-3]
senators.o[[1]] = senators.o[[1]][,-3]
senators.p[[1]] = senators.p[[1]][,-3]
senators.q[[1]] = senators.q[[1]][,-3]
senators.r[[1]] = senators.r[[1]][,-3]
senators.s[[1]] = senators.s[[1]][,-3]
senators.t[[1]] = senators.t[[1]][,-3]
senators.u[[1]] = senators.u[[1]][,-3]
senators.v[[1]] = senators.v[[1]][,-3]
senators.w[[1]] = senators.w[[1]][,-3]
senators.y[[1]] = senators.y[[1]][,-3]
senators.z[[1]] = senators.z[[1]][,-3]
```


### Next, bind the scraped lists into a single dataframe

```{r eval=TRUE}
#rbind is easiest solution for this..
us.senators <- rbind(senators.a[[1]],
                     senators.b[[1]],
                     senators.c[[1]],
                     senators.d[[1]],
                     senators.e[[1]],
                     senators.f[[1]],
                     senators.g[[1]],
                     senators.h[[1]],
                     senators.i[[1]],
                     senators.j[[1]],
                     senators.k[[1]],
                     senators.l[[1]],
                     senators.m[[1]],
                     senators.n[[1]],
                     senators.o[[1]],
                     senators.p[[1]],
                     senators.q[[1]],
                     senators.r[[1]],
                     senators.s[[1]],
                     senators.t[[1]],
                     senators.u[[1]],
                     senators.v[[1]],
                     senators.w[[1]],
                     senators.y[[1]],
                     senators.z[[1]])


head(us.senators)
```

## Now, fix the one row that is formatted
## differently than all the others (row 1112)

```{r}

us.senators[1112,2]
us.senators[1112,2] <- '2005-2009'
us.senators[1112,2]

#for (i in 1:1889){
#  str_replace(us.senators$Years[[i]], '2005-2009', '2005-2009')
#}
```


# OK, the current dataframe has all of our wanted
# information but it is not clean nor tidy..

## So, let's tidy some the data for aggregation

```{r}
# first, import the stringr package
library(stringr)

# Some senators had gaps of time between their
# stints in the senate. Thus, we must first break
# their time in the senate into time spans:

# added difficulty: some senators took over mid-term..


# So, in the following loops we are breaking the Year
# column into multiple columns for each set of years a Senator was
# in office. A senator may have multiple sets of years for two reasons:
# 1. they switched parties
# 2. they had a gap from their time in the senate

for (i in 1:1889){
  if (str_sub(us.senators$Years[[i]], 5, 5) == '-'){
    us.senators$first.years[[i]] <- str_sub(us.senators$Years[[i]], 1, 9)
    us.senators$Years[[i]] <- str_sub(us.senators$Years[[i]], 10, str_length(us.senators$Years[[i]]))
  } else {
    us.senators$first.years[[i]] <- str_sub(us.senators$Years[[i]], 1, 4)
    us.senators$Years[[i]] <- str_sub(us.senators$Years[[i]], 5, str_length(us.senators$Years[[i]]))
  }
}

for (i in 1:1889){
  if (str_sub(us.senators$Years[[i]], 5, 5) == '-'){
    us.senators$second.years[[i]] <- str_sub(us.senators$Years[[i]], 1, 9)
    us.senators$Years[[i]] <- str_sub(us.senators$Years[[i]], 10, str_length(us.senators$Years[[i]]))
  } else {
    us.senators$second.years[[i]] <- str_sub(us.senators$Years[[i]], 1, 4)
    us.senators$Years[[i]] <- str_sub(us.senators$Years[[i]], 5, str_length(us.senators$Years[[i]]))
  }
}

for (i in 1:1889){
  if (str_sub(us.senators$Years[[i]], 5, 5) == '-'){
    us.senators$third.years[[i]] <- str_sub(us.senators$Years[[i]], 1, 9)
    us.senators$Years[[i]] <- str_sub(us.senators$Years[[i]], 10, str_length(us.senators$Years[[i]]))
  } else {
    us.senators$third.years[[i]] <- str_sub(us.senators$Years[[i]], 1, 4)
    us.senators$Years[[i]] <- str_sub(us.senators$Years[[i]], 5, str_length(us.senators$Years[[i]]))
  }
}

for (i in 1:1889){
  if (str_sub(us.senators$Years[[i]], 5, 5) == '-'){
    us.senators$fourth.years[[i]] <- str_sub(us.senators$Years[[i]], 1, 9)
    us.senators$Years[[i]] <- str_sub(us.senators$Years[[i]], 10, str_length(us.senators$Years[[i]]))
  } else {
    us.senators$fourth.years[[i]] <- str_sub(us.senators$Years[[i]], 1, 4)
    us.senators$Years[[i]] <- str_sub(us.senators$Years[[i]], 5, str_length(us.senators$Years[[i]]))
  }
}

for (i in 1:1889){
  if (str_sub(us.senators$Years[[i]], 5, 5) == '-'){
    us.senators$fifth.years[[i]] <- str_sub(us.senators$Years[[i]], 1, 9)
    us.senators$Years[[i]] <- str_sub(us.senators$Years[[i]], 10, str_length(us.senators$Years[[i]]))
  } else {
    us.senators$fifth.years[[i]] <- str_sub(us.senators$Years[[i]], 1, 4)
    us.senators$Years[[i]] <- str_sub(us.senators$Years[[i]], 5, str_length(us.senators$Years[[i]]))
  }
}
```

### This following chunk is simple, it's filling
### the empty (new) columns with zero's
```{r}
# Now, to fill in all the blank cells with zero's
for (i in 7:10){
  for (j in 1:1889){
    if (us.senators[j, i] == ""){
      us.senators[j,i] <- 0
    }
  }
}
```


## Now, we'll start our long process to identify the specific
## years each senator was in office, not just the ranges

```{r}
# Split each set of years into two columns

us.senators$first.years.first <- str_sub(us.senators$first.years, 1, 4)
us.senators$first.years.last <- str_sub(us.senators$first.years, 6, 10)

us.senators$second.years.first <- str_sub(us.senators$second.years, 1, 4)
us.senators$second.years.last <- str_sub(us.senators$second.years, 6, 10)

us.senators$third.years.first <- str_sub(us.senators$third.years, 1, 4)
us.senators$third.years.last <- str_sub(us.senators$third.years, 6, 10)

us.senators$fourth.years.first <- str_sub(us.senators$fourth.years, 1, 4)
us.senators$fourth.years.last <- str_sub(us.senators$fourth.years, 6, 10)

us.senators$fifth.years.first <- str_sub(us.senators$fifth.years, 1, 4)
us.senators$fifth.years.last <- str_sub(us.senators$fifth.years, 6, 10)

# again, fill in the blank cells with zero's

for (i in 11:20){
  for (j in 1:1889){
    if (us.senators[j, i] == ""){
      us.senators[j,i] <- 0
    } else {
      us.senators[j,i] <- as.numeric(us.senators[j,i])
    }
  }
}

for (i in 11:20){
  for (j in 1:1889){
    us.senators[j,i] <- as.numeric(us.senators[j,i])
  }
}
```

us.senators[100,11]

```{r warning=FALSE}
# create lists of years, ranges, for each
# set of years a senator was in office

# this is all still to achieve a goal of
# one row for each year a senator was in office

# the hard part is each row much correspond
# to the political party they were part of at
# that time..

us.senators$first.years.range <- list(0)
us.senators$second.years.range <- list(0)
us.senators$third.years.range <- list(0)
us.senators$fourth.years.range <- list(0)
us.senators$fifth.years.range <- list(0)

for (i in 1:1889){
  if (us.senators$first.years.first[[i]] != 0 &
    us.senators$first.years.last[[i]] != 0) {
    us.senators$first.years.range[[i]] <- seq(us.senators$first.years.first[[i]],
                                              as.numeric(us.senators$first.years.last[[i]])-1, by=1)
  } else {
    us.senators$first.years.range[[i]] <- list(us.senators$first.years.first[[i]])
  }
}

for (i in 1:1889){
  if (us.senators$second.years.first[[i]] != 0 &
    us.senators$second.years.last[[i]] != 0) {
    us.senators$second.years.range[[i]] <- seq(us.senators$second.years.first[[i]],
                                              as.numeric(us.senators$second.years.last[[i]])-1, by=1)
  } else {
    us.senators$second.years.range[[i]] <- list(us.senators$second.years.first[[i]])
  }
}

for (i in 1:1889){
  if (us.senators$third.years.first[[i]] != 0 &
    us.senators$third.years.last[[i]] != 0) {
    us.senators$third.years.range[[i]] <- seq(us.senators$third.years.first[[i]],
                                              as.numeric(us.senators$third.years.last[[i]])-1, by=1)
  } else {
    us.senators$third.years.range[[i]] <- list(us.senators$third.years.first[[i]])
  }
}

for (i in 1:1889){
  if (us.senators$fourth.years.first[[i]] != 0 &
    us.senators$fourth.years.last[[i]] != 0) {
    us.senators$fourth.years.range[[i]] <- seq(us.senators$fourth.years.first[[i]],
                                              as.numeric(us.senators$fourth.years.last[[i]])-1, by=1)
  } else {
    us.senators$fourth.years.range[[i]] <- list(us.senators$fourth.years.first[[i]])
  }
}

for (i in 1:1889){
  if (us.senators$fifth.years.first[[i]] != 0 &
    us.senators$fifth.years.last[[i]] != 0) {
    us.senators$fifth.years.range[[i]] <- seq(us.senators$fifth.years.first[[i]],
                                              as.numeric(us.senators$fifth.years.last[[i]])-1, by=1)
  } else {
    us.senators$fifth.years.range[[i]] <- list(us.senators$fifth.years.first[[i]])
  }
}
```

## From the above few chunks we sucessfully split the original messy
## Year column into lists of years each senator was in office

## Now, we must tie each year a senators was in office to the political
## party they represented during that year

```{r warning=FALSE}
# Now an additional challenge, some senators served
# their time in senate under multiple political parties.
# To address this, let's split each senators time in
# office by their political party

# first, create a list of all the parties

pol.parties <- paste(c("Democratic-Republican",
                      "Federalist", "Whig", "National Republican",
                      "Populist", "Liberal Republican", "Unionist",
                      "Farmer-Laborite", "Free Soiler", "Readjuster",
                      "Independant", "Progressive", "Silver Republican",
                      "American", "Anti-Monopolist", "Conservative", "Independent",
                      "Nullifier", "Independant Democrat", "Independence",
                      "Law and Order", "Liberty", "Oppositionist", "Silverite",
                      "Unconditional Unionist", "Democratic", "Republican", '"'), collapse = "|")

us.senators$political.parties <- str_extract_all(us.senators$Party, pol.parties)

for (i in 1:1889){
  us.senators$first.party[[i]] <- us.senators$political.parties[[i]][1]
  us.senators$second.party[[i]] <- us.senators$political.parties[[i]][2]
  us.senators$third.party[[i]] <- us.senators$political.parties[[i]][3]
  us.senators$fourth.party[[i]] <- us.senators$political.parties[[i]][4]
  us.senators$fifth.party[[i]] <- us.senators$political.parties[[i]][5]
}


# But as you may notice, wikipedia used " to show the same political
# party as the previous set of years. Thus, we need to change all the
# cells that have "

for (i in 1:1889){
  if (!is.na(us.senators$second.party[[i]]) & us.senators$second.party[[i]] == '\"'){
    us.senators$second.party[[i]] <- us.senators$first.party[[i]]
  } else {}
}

for (i in 1:1889){
  if (!is.na(us.senators$third.party[[i]]) &
      us.senators$third.party[[i]] == '\"' &
      us.senators$second.party[[i]] != '\"'){
    us.senators$third.party[[i]] <- us.senators$second.party[[i]]
  } else if (!is.na(us.senators$third.party[[i]]) &
      us.senators$third.party[[i]] == '\"' &
      us.senators$second.party[[i]] == '\"'){
    us.senators$third.party[[i]] <- us.senators$first.party[[i]]
  }
}

for (i in 1:1889){
  if (!is.na(us.senators$fourth.party[[i]]) &
      us.senators$fourth.party[[i]] == '\"' &
      us.senators$third.party[[i]] != '\"'){
    us.senators$fourth.party[[i]] <- us.senators$third.party[[i]]
  } else if (!is.na(us.senators$fourth.party[[i]]) &
      us.senators$fourth.party[[i]] == '\"' &
      us.senators$third.party[[i]] == '\"'){
    us.senators$fourth.party[[i]] <- us.senators$second.party[[i]]
  }
}

for (i in 1:1889){
  if (!is.na(us.senators$fifth.party[[i]]) &
      us.senators$fifth.party[[i]] == '\"' &
      us.senators$fourth.party[[i]] != '\"'){
    us.senators$fifth.party[[i]] <- us.senators$fourth.party[[i]]
  } else if (!is.na(us.senators$fifth.party[[i]]) &
      us.senators$fifth.party[[i]] == '\"' &
      us.senators$fourth.party[[i]] == '\"'){
    us.senators$fifth.party[[i]] <- us.senators$third.party[[i]]
  }
}


```

```{r warning=FALSE}
# one catch, if a senator has two seperate ranges of years,
# but, they did not switch parties, we have elminiated their
# party in the second range. to adjust that:

for (i in 1:1889){
  if (length(us.senators$second.years.range[[i]])>1 &
      (us.senators$second.years.range[[i]][1])!= "0" &
      is.na(us.senators$second.party) == TRUE) {
    us.senators$second.party[[i]] <- us.senators$first.party[[i]]
  }
}
```


## Finally, we have all the information we need!
## So, let's delete the unneccsary columns

```{r}
library(dplyr)

us.senators <- select(us.senators, -c(2, 5, 6, 7, 8,
                                      9, 10, 11, 12,
                                      13, 14, 15, 16,
                                      17, 18, 19, 20))
```


# One last thing: we need to add all the current senators!
# In our dataframe now, we only have previous senators
```{r}
add.current.senators <- read_csv("data/add_current_senators.csv")
head(add.current.senators)
```

### format the current sentaors to merge with us.senators dataframe

```{r}
add.current.senators$first.years.range.end <- 2019
add.current.senators$first.years.range <- as.numeric(add.current.senators$first.years.range)
add.current.senators$first.years.range.two <- list(0)

for (i in 1:100){
  add.current.senators$first.years.range.two[[i]] <- seq(add.current.senators$first.years.range[[i]],
                                              add.current.senators$first.years.range.end[[i]],
                                              by=1)
}

add.current.senators$first.years.range <- add.current.senators$first.years.range.two

for (i in 1:100){
  add.current.senators$second.years.range[[i]] <- list(0)
  add.current.senators$third.years.range[[i]] <- list(0)
  add.current.senators$fourth.years.range[[i]] <- list(0)
  add.current.senators$fifth.years.range[[i]] <- list(0) 
}


add.current.senators <- select(add.current.senators, -c(15,16))

names(add.current.senators)
```

# add the current senators into the dataframe with all previous senators

```{r}
us.senators.final <- rbind(us.senators, add.current.senators)
```



## To tranform the dataframe to have one row for each
## year per senator

```{r warning=FALSE}
senator.by.year <- data.frame("Senator" = 0, "Year" = 0, "State" = 0, "Party"= 0)

# Summary: the following loop creates a single row entry
# for each year of each senator's time in office

for (i in 1:1989){
  for (j in 1:length(us.senators.final$first.years.range[[i]])){
    #print(us.senators$Senator[[i]])
    #print(us.senators$first.years.range[[i]][[j]])
    #print(us.senators$State[[i]])
    #print(us.senators$Party[[i]])
    nextrow = list(us.senators.final$Senator[[i]],
                   us.senators.final$first.years.range[[i]][[j]],
                   us.senators.final$State[[i]],
                   us.senators.final$first.party[[i]])
    senator.by.year <- rbind(senator.by.year, nextrow)
  }
}

# drop the first row, it was merely a test
senator.by.year <- senator.by.year[-1,]

```

```{r warning=FALSE}
# run the same loop for the second set of years
for (i in 1:1989){
  for (j in 1:length(us.senators.final$second.years.range[[i]])){
    if (us.senators.final$second.years.range[[i]] > 0){
      nextrow = list(us.senators.final$Senator[[i]],
                     us.senators.final$second.years.range[[i]][[j]],
                     us.senators.final$State[[i]],
                     us.senators.final$second.party[[i]])
      senator.by.year <- rbind(senator.by.year, nextrow)
    }
  }
}
```

```{r warning=FALSE}
# run the same loop for the third set of years
for (i in 1:1989){
  for (j in 1:length(us.senators.final$third.years.range[[i]])){
    if (us.senators.final$third.years.range[[i]] > 0){
      nextrow = list(us.senators.final$Senator[[i]],
                     us.senators.final$third.years.range[[i]][[j]],
                     us.senators.final$State[[i]],
                     us.senators.final$third.party[[i]])
      senator.by.year <- rbind(senator.by.year, nextrow)
    }
  }
} 
```

```{r warning=FALSE}
# run the same loop for the fourth set of years
for (i in 1:1989){
  for (j in 1:length(us.senators.final$fourth.years.range[[i]])){
    if (us.senators.final$fourth.years.range[[i]] > 0){
      nextrow = list(us.senators.final$Senator[[i]],
                     us.senators.final$fourth.years.range[[i]][[j]],
                     us.senators.final$State[[i]],
                     us.senators.final$fourth.party[[i]])
      senator.by.year <- rbind(senator.by.year, nextrow)
    }
  }
} 

```

```{r warning=FALSE}
# run the same loop for the fifth set of years
for (i in 1:1989){
  for (j in 1:length(us.senators.final$fifth.years.range[[i]])){
    if (us.senators.final$fifth.years.range[[i]] > 0){
      nextrow = list(us.senators.final$Senator[[i]],
                     us.senators.final$fifth.years.range[[i]][[j]],
                     us.senators.final$State[[i]],
                     us.senators.final$fifth.party[[i]])
      senator.by.year <- rbind(senator.by.year, nextrow)
    }
  }
} 
```


# Ok, over 600 lines of code to organize the senate data!
## Now, let's do some quick checks on the data

```{r}
library(tidyr)

senator.by.year %>%
  group_by(Year) %>%
  summarise(Count = n()) %>%
  arrange(desc(Year))
```

```{r}
write.csv(senator.by.year, "Senator_by_year.csv")

us.senators.final <- select(us.senators.final, c(1, 2, 10))
write.csv(us.senators.final, "All_Senators.csv")
```

```{r}
us.senators.final %>%
  group_by(first.party) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))
```
```{r}
senator.by.year %>%
  group_by(Party) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))
```

```{r warning=FALSE}
filter(senator.by.year, State == "Georgia") %>%
  group_by(Party) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))
```
